Information in recombination junctions can both compress and improve inferred tree sequences

### 2023-10-02 paper/writeup.tex
template; start at abstract/intro

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usetikzlibrary{arrows, snakes,backgrounds}
\tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}

@@

@@

The pedigree that describes how everyone of some species
are related to each other
is a large graph, where nodes are indexed by time.
Within this, one can trace back the paths along which segement of genome
have been inherited.
The \emph{succinct tree sequence} (or, tree sequence for short)
was introduced by \citet{kelleher2016efficient} to describe the subset of these paths
along which the genome of a sample of individuals was inherited.

@@

\begin{algorithm}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??

@@

\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
{\BlankLine
	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
}
\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
	\For{e1 \KwTo EdgesOut}{
		\For{e2 \KwTo EdgesOut}{
			\If{e2.parent = e1.child}{
				\For{e3 \KwTo EdgesIn}{
					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
						e1.right $\gets$ Interval.right\;
						e2.right $\gets$ Interval.right\;
						e3.left $\gets$ Interval.right\;
						\If{e3.left = e3.right}{
							$E$.remove(e3)\;
							}}}}}}}\;
$TS'\gets TS$.update()\;						
\KwRet{$TS'$}\;
}
\BlankLine						
\Fn{\ex ($TS$, iter=100)}{
	edges $\gets$ $TS$.numedges()\;
	\BlankLine
	\For{$*$ \KwTo \Range(iter)}{
		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
		\eIf{$TS$.numedges = edges}{
			\Break\BlankLine}
		{edges $\gets$ $TS'$.numedges()\;}
		}
}
\KwRet{$TS'$}\;
\end{algorithm}

@@

\begin{algorithm}[!ht]
\SetStartEndCondition{ }{}{}%
\SetKwProg{Fn}{def}{\string:}{}
\SetKwFunction{Range}{range}%%
\SetKw{KwTo}{in}
\SetKwFor{For}{for}{\string:}{}%
\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
\SetKwFor{While}{while}{:}{fintq}%
\SetKw{Break}{break}
\newcommand\forcond{$i$ \KwTo\Range{$n$}}
\SetKwComment{Comment}{}{} 
\DontPrintSemicolon
\LinesNotNumbered
\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%

\input{$(TS_1,TS_2)$}
\output{discrepancy number (some float value)}

\Fn{Node Discrepancy($x,TS_1,TS_2$)}{
$s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}

\For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
\Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
$D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
\If{$|D_{T^1_k}(x)| > 0$}{
\If{$|D_{T^1_k}(x)|=1$}{
$y \gets D_{T^1_k}(x)$\;
$m \gets D_{T^1_k}(x)$\;}
\Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
$y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
\If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
$s_y\gets s_y$\;}}
\KwRet{node discrepancy}\;}
\Fn{Discrepancy($TS_1,TS_2$)}{
discrepancy = 0\;
\For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
\BlankLine
\KwRet{discrepancy}}
\caption{Discrepancy Function}\label{ndisc}
\end{algorithm}

>>>>

\documentclass[10pt,twoside,lineno]{gsajnl}
\articletype{inv} % article type

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined,algo2e]{algorithm2e}
\usepackage{tikz}
\usetikzlibrary{arrows, snakes,backgrounds}
\tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]

\bibliographystyle{plainnat}

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}

\title{
    Information in recombination junctions 
    can both compress and improve inferred tree sequences
}

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

\affil[$\ast$]{Institute of Evolution and Ecology and Department of Biology, University of Oregon, Eugene, Oregon}
\affil[$\dagger$]{Department of Mathematics, University of Oregon, Eugene, Oregon}
\affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}

\keywords{genealogy, tree sequence, haplotypes}

\runningtitle{Junctions in tree sequences}
\runningauthor{Bamforth \textit{et al.}}

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on
    polymorphic markers (i.e., information produced by mutation),
    which produces unary nodes along the branches of the marginal genealogical trees.
    The resulting inferred ARGs, represented as tree sequences,
    are smaller, faster to compute with,
    and potentially contain substantially more information
    about the age of ancestral haplotypes,
    than inferred ARGs without these unary nodes.
    We provide efficient algorithms to identify this ancestral haplotype information,
    new metrics of agreement/disagreement between inferred ARGs,
    and explore some consequences for ARGs inferred from real data.
\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{1}{Corresponding author: {plr@uoregon.edu}}
\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page

@@

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome.
For reviews of this progress, see CITE CITE CITE.
Often such methods focus on the sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
For instance, the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit (CITE).

However, the ARG is emphatically not merely a sequence of trees:
since each segment of an ancestor's genome that has been inherited by the focal genomes
may extend over a portion of the genome over which there are many distinct trees,
a single ancestor should usually be represented by nodes in many distinct trees.
In other words, one way to view an ARG is as a sequence of trees in which nodes may be shared across many trees,
but another way to view the ARG is as a set of relationships between segments of ancestral haplotypes.

Another reason we tend to focus on the trees is that
much of our intuition about inference of relationships from genomic data
comes from phylogenetics.
Indeed, all methods might very roughly be summarized as
``more similar sequences are more closely related''.
For instance, two sequences that share a drerived mutation
are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
\citet{fisher} pointed out that not only mutations
but also the ``junctions'' between distinct haplotypes,
if they could be somehow identified,
would be inherited as Mendelian markers.

@@

\begin{algorithm2e}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??

@@

\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
{\BlankLine
	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
}
\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
	\For{e1 \KwTo EdgesOut}{
		\For{e2 \KwTo EdgesOut}{
			\If{e2.parent = e1.child}{
				\For{e3 \KwTo EdgesIn}{
					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
						e1.right $\gets$ Interval.right\;
						e2.right $\gets$ Interval.right\;
						e3.left $\gets$ Interval.right\;
						\If{e3.left = e3.right}{
							$E$.remove(e3)\;
							}}}}}}}\;
$TS'\gets TS$.update()\;						
\KwRet{$TS'$}\;
}
\BlankLine						
\Fn{\ex ($TS$, iter=100)}{
	edges $\gets$ $TS$.numedges()\;
	\BlankLine
	\For{$*$ \KwTo \Range(iter)}{
		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
		\eIf{$TS$.numedges = edges}{
			\Break\BlankLine}
		{edges $\gets$ $TS'$.numedges()\;}
		}
}
\KwRet{$TS'$}\;
\end{algorithm2e}

@@

% PLR: commented this out since it's not compiling
% 
% \begin{algorithm}[!ht]
% \SetStartEndCondition{ }{}{}%
% \SetKwProg{Fn}{def}{\string:}{}
% \SetKwFunction{Range}{range}%%
% \SetKw{KwTo}{in}
% \SetKwFor{For}{for}{\string:}{}%
% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
% \SetKwFor{While}{while}{:}{fintq}%
% \SetKw{Break}{break}
% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
% \SetKwComment{Comment}{}{} 
% \DontPrintSemicolon
% \LinesNotNumbered
% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
% 
% 
% \input{$(TS_1,TS_2)$}
% \output{discrepancy number (some float value)}
% 
% \Fn{Node Discrepancy($x,TS_1,TS_2$)}{
% $s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}
% 
% \For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
% \Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
% $D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
% \If{$|D_{T^1_k}(x)| > 0$}{
% \If{$|D_{T^1_k}(x)|=1$}{
% $y \gets D_{T^1_k}(x)$\;
% $m \gets D_{T^1_k}(x)$\;}
% \Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
% $y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
% \If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
% $s_y\gets s_y$\;}}
% \KwRet{node discrepancy}\;}
% \Fn{Discrepancy($TS_1,TS_2$)}{
% discrepancy = 0\;
% \For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
% \BlankLine
% \KwRet{discrepancy}}
% \caption{Discrepancy Function}\label{ndisc}
% \end{algorithm}
